---
title: "ConvergentMCMC_SIR"
author: "Brian Masafu"
date: "2024-09-26"
output:
  pdf_document: default
  html_document: default
---
# SIR_CompartmentalWith_3_AgeStructuredComparments 
# libraries
```{r}
library(deSolve)
library(tidyverse)
library(coda)
library(MCMCvis)
library(lattice)

```


# SIR model function with age compartments
```{r}
SIR_model_age <- function(time, state, parameters) {
  # Children (age group 1)
  S1 <- state[1]
  I1 <- state[2]
  R1 <- state[3]
  
  # Adults (age group 2)
  S2 <- state[4]
  I2 <- state[5]
  R2 <- state[6]
  
  # Elderly (age group 3)
  S3 <- state[7]
  I3 <- state[8]
  R3 <- state[9]
  
  # Total population for each age group
  N1 <- S1 + I1 + R1
  N2 <- S2 + I2 + R2
  N3 <- S3 + I3 + R3
  
  # Transmission rates (force of infection) for each age group
  beta1 <- parameters["beta1"]
  beta2 <- parameters["beta2"]
  beta3 <- parameters["beta3"]
  gamma <- parameters["gamma"]
  
  # Equations for age group 1 (Children)
  dS1 <- -beta1 * S1 * I1 / N1 
  dI1 <- beta1 * S1 * I1 / N1 - gamma * I1
  dR1 <- gamma * I1
  
  # Equations for age group 2 (Adults)
  dS2 <- -beta2 * S2 * I2 / N2
  dI2 <- beta2 * S2 * I2 / N2 - gamma * I2
  dR2 <- gamma * I2
  
  # Equations for age group 3 (Elderly)
  dS3 <- -beta3 * S3 * I3 / N3
  dI3 <- beta3 * S3 * I3 / N3 - gamma * I3
  dR3 <- gamma * I3
  
  # Returning the rates of change for all compartments
  list(c(dS1, dI1, dR1, dS2, dI2, dR2, dS3, dI3, dR3))
}

```

# Parameters

```{r}
beta1 <- 0.4   # Force of infection for children
beta2 <- 0.3   # Force of infection for adults
beta3 <- 0.25  # Force of infection for elderly
gamma <- 0.1 
```

# Initial conditions for each age group &time sequence
```{r}
# Initial conditions for each age group
initial_state <- c(S1 = 50, I1 = 1, R1 = 0,  # Children
                   S2 = 40, I2 = 1, R2 = 0,  # Adults
                   S3 = 30, I3 = 1, R3 = 0)  # Elderly

# Time sequence (one year with daily time steps)
times <- seq(0, 365, by = 1)
```

# simulation of the model & plotting of the three states
```{r}
# Simulation of the SIR model with age structure
parameters <- c(beta1 = beta1, beta2 = beta2, beta3 = beta3, gamma = gamma)
simulated_data <- ode(y = initial_state, times = times, func = SIR_model_age, parms = parameters) %>% data.frame()

# Ensuring that the observed data columns are numeric
simulated_data[, 2:10] <- lapply(simulated_data[, 2:10], as.numeric)

# Adding a Time column for plotting
simulated_data$Time <- times

# Reshaping data to long format for plotting
simulated_data_long <- simulated_data %>%
  pivot_longer(cols = -Time, names_to = "Compartment", values_to = "Count")

# Plotting all states on one graph with correct color mapping and y-axis limits
ggplot(simulated_data_long, aes(x = Time, y = Count, color = Compartment)) +
  geom_line() +
  labs(title = "SIR Model Simulation with 3_Agecompartments", x = "Time (days)", y = "Count") +
  scale_color_manual(values = c("blue", "green", "red",      # Children
                                "blue4", "green4", "red4",  # Adults
                                "blue2", "green2", "red2",  # Elderly
                                "black")) +                  # Recovered (R1, R2, R3)
  coord_cartesian(ylim = c(0, 100)) +  #  y-axis limits from 0 to 100
  theme_minimal()

```
SIR_plots showing different levels of force of infections and recovery rates among the 3 age structured population with (Susceptible, Infected and Recovered States)

# Log-likelihood function for the model with age compartments
```{r}
log_likelihood_age <- function(params, observed_data, initial_state, times) {
  
  # Parameters with the current estimate of beta for each age group
  parameters <- c(params["beta1"], params["beta2"], params["beta3"], gamma = 1/10)
  
  # Simulating the model
  out <- ode(y = initial_state, times = times, func = SIR_model_age, parms = parameters)
  model_data <- as.data.frame(out)
  
  #checking observed data(numeric form)
  observed_data[, 2:10] <- lapply(observed_data[, 2:10], as.numeric)
  
  # Log-likelihood for each compartment in each age group
  log_likelihood <- 0
  for (col in 2:10) {
    obs <- ceiling(observed_data[[col]])
    model <- model_data[[col]]
    log_likelihood <- log_likelihood + sum(dpois(obs, lambda = model, log = TRUE))
  }
  
  return(log_likelihood)
}
```

# MCMC settings
```{r}

n_iter <- 5000
beta_init <- c(0.1, 0.15, 0.2)  # Initial beta values for MCMC
n_chains <- 3
sd_prop <- rep(0.001, n_chains)  # Initial SD for each chain
target_accept_rate <- 0.234  # Target acceptance rate for adaptive MCMC
adapt_rate <- 0.01  # Rate of adaptation for the proposal SD
```

# Incorporating RHAT logic: storage for multiple chains& initializing the chain
```{r}
# Incorporating RHAT logic: storage for multiple chains
beta_chains <- matrix(NA, ncol = n_chains, nrow = n_iter)

# Initializing chains with different initial values
beta_chains[1, ] <- beta_init

# Prior distribution: Beta(2, 2)
prior <- function(beta) {
  return(dbeta(beta, 2, 2, log = TRUE))
}
```
Rhat(potential scale reduction factor) - Measures whether or not an MCMC algorithm converged. It checks the distribution of a chain (after warm up ) the same as the distribution of the second half of the chain.
2. If the algorithm starts at 2 different places and chain left to warm up, both the chains have the same distribution.


# MCMC loop for each chain
```{r}

# MCMC loop for each chain
for (chain in 1:n_chains) {
  
  acceptance_counter <- 0 # Reset acceptance counter for each chain
  
  # Initializing log likelihood for current chain
  loglik_curr <- log_likelihood_age(params = c(beta1 = beta_chains[1, chain], beta2 = beta_chains[1, chain], beta3 = beta_chains[1, chain]),
                                    observed_data = simulated_data,
                                    initial_state = initial_state,
                                    times = times) +
    prior(beta_chains[1, chain])
  
  for (i in 2:n_iter) {
    # Proposing new beta from a normal distribution centered around the current value
    beta_proposed <- rnorm(1, mean = beta_chains[i - 1, chain], sd = sd_prop[chain])
    
    if (beta_proposed > 0) {  # Ensure beta is positive
      loglik_prop <- log_likelihood_age(params = c(beta1 = beta_proposed, beta2 = beta_proposed, beta3 = beta_proposed),
                                        observed_data = simulated_data,
                                        initial_state = initial_state,
                                        times = times) + 
        prior(beta_proposed)
    } else {
      loglik_prop <- -1E6  # Penalize invalid proposals
    }
    
    # Calculating acceptance probability
    acceptance_prob <- loglik_prop - loglik_curr
    
    # Metropolis-Hastings acceptance step
    if (log(runif(1)) < acceptance_prob) {
      beta_chains[i, chain] <- beta_proposed
      loglik_curr <- loglik_prop
      acceptance_counter <- acceptance_counter + 1
    } else {
      beta_chains[i, chain] <- beta_chains[i - 1, chain]
    }
    
    # Adaptive adjustment of proposal standard deviation (sd_prop)
    if (i > 100) {  # Adapt every 100 iterations
      acceptance_rate <- acceptance_counter / i
      sd_prop[chain] <- sd_prop[chain] * exp(adapt_rate * (acceptance_rate - target_accept_rate))
    }
  }
}

```

#Assigning parameter names and converting the chains into compartible data
```{r}
# Assigning parameter names to the columns in beta_chains
colnames(beta_chains) <- rep("Beta", n_chains)

#  mcmc.list 
burnin <- 1000
mcmc_out <- mcmc.list(
  as.mcmc(beta_chains[burnin:n_iter, 1, drop = FALSE]),
  as.mcmc(beta_chains[burnin:n_iter, 2, drop = FALSE]),
  as.mcmc(beta_chains[burnin:n_iter, 3, drop = FALSE])
)

# Converting beta_chains matrix into mcmc objects 
mcmc_chain1 <- as.mcmc(beta_chains[, 1])
mcmc_chain2 <- as.mcmc(beta_chains[, 2])
mcmc_chain3 <- as.mcmc(beta_chains[, 3])

# Combining all chains into an mcmc.list object
mcmc_combined <- mcmc.list(mcmc_chain1, mcmc_chain2, mcmc_chain3)

# Calculating R-hat using rs tan's summary function
rhat_values <- rstan::Rhat(as.matrix(mcmc_combined))


```

# Printing R-hat values for convergence diagnostics

```{r}
# Printing R-hat values for convergence diagnostics
print(rhat_values)

```
Rhat values close to 1(1.1) indicates that the chain has converged. Values greater than 1.1 suggests that the chain has not yet converged and you may need to run more iteration. `Rhat calculations` You need multiple chains to calculate Rhat.
#plots for trace mcmc
```{r}

#  trace plot using coda's plot function
plot(mcmc_combined, trace = TRUE, density = FALSE, main = "Trace Plots for MCMC Chains")

# Assigning column names to the beta_chains matrix 
colnames(beta_chains) <- c("Beta1", "Beta2", "Beta3")

# Converting each chain to mcmc objects again
mcmc_chain1 <- as.mcmc(matrix(beta_chains[, "Beta1"], ncol = 1, dimnames = list(NULL, "Beta")))
mcmc_chain2 <- as.mcmc(matrix(beta_chains[, "Beta2"], ncol = 1, dimnames = list(NULL, "Beta")))
mcmc_chain3 <- as.mcmc(matrix(beta_chains[, "Beta3"], ncol = 1, dimnames = list(NULL, "Beta")))

# Combining  into an mcmc.list object
mcmc_combined <- mcmc.list(mcmc_chain1, mcmc_chain2, mcmc_chain3)

# MCMCtrace 
MCMCtrace(mcmc_combined, params = "Beta", pdf = FALSE)

# Ploting posterior density using coda's densityplot function
densityplot(mcmc_combined, main = "Posterior Density of Beta")


# posterior  visualization using MCMCvis
MCMCplot(mcmc_combined, params = "Beta", main = "Posterior Distribution of Beta")

```

# Calculate Gelman and Rubin's diagnostic (R-hat)
```{r}
gelman.diag(mcmc_combined)
```
This MCMC chains have likely converged to the same target distribution, and there is little difference between the chains' behavior.Values close to 1 (typically less than 1.1) are considered evidence of convergence.

# MCMCvis to summary of chains including R-hat and effective sample size
```{r}

MCMCsummary(mcmc_combined, Rhat = TRUE, n.eff = TRUE)
```

# rstan for comprehensive summary including n_eff and other statistics
```{r}

rstan::summary(mcmc_combined)
```

# MCMC Trace for Beta && # Histogram of posterior distribution
```{r}
# MCMC Trace for Beta
mcmc_df <- data.frame(Iteration = burnin:n_iter, Beta = c(beta_chains[burnin:n_iter, ]))
ggplot(mcmc_df, aes(x = Iteration, y = Beta)) +
  geom_line(color = "blue", linewidth = 1) +
  labs(title = "MCMC Trace for Beta (Adaptive MCMC)", x = "Iteration", y = "Beta") +
  theme_minimal()

# Histogram of posterior distribution
ggplot(data.frame(Beta = c(beta_chains[burnin:n_iter, ])), aes(x = Beta)) +
  geom_histogram(bins = 50, fill = "skyblue", color = "black") +
  labs(title = "Posterior Distribution of Beta", x = "Beta", y = "Frequency") +
  theme_minimal()

```